---
title: "Engy Fouda assign. 11"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
library(ISLR)
library(e1071)
library(randomForest)
library(class)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
```


# Preface

This week assignment will explore behavior of support vector classifiers and SVMs (following the distinction made in ISLR) on banknote authentication dataset from UCI ML archive.  We worked with it on multiple occasions before (most recently two weeks ago evaluating performance of logistic regression, discriminant analysis and KNN on it):

```{r dbaExample}
dbaDat <- read.table("C:/Users/Engy/Downloads/Harvard/7th semester ISA/R/Week 11/data_banknote_authentication.txt",sep=",")
colnames(dbaDat) <- c("var","skew","curt","entr","auth")
dbaDat$auth <- factor(dbaDat$auth)
dim(dbaDat)
summary(dbaDat)
head(dbaDat)
pairs(dbaDat[,1:4],col=as.numeric(dbaDat$auth))
```

Here we will use SVM implementation available in library `e1071` to fit classifiers with linear and radial (polynomial for extra points) kernels and compare their relative performance as well as to that of random forest and KNN.

# Problem 1 (20 points): support vector classifier (i.e. using linear kernel) 

Use `svm` from library `e1071` with `kernel="linear"` to fit classifier (e.g. ISLR Ch.9.6.1) to the entire banknote authentication dataset setting parameter `cost` to 0.001, 1, 1000 and 1 mln.  Describe how this change in parameter `cost` affects model fitting process (hint: the difficulty of the underlying optimization problem increases with cost -- can you explain what drives it?) and its outcome (how does the number of support vectors change with `cost`?) and what are the implications of that.  Explain why change in `cost` value impacts number of support vectors found. (Hint: there is an answer in ISLR.)  Use `tune` function from library `e1071` (see ISLR Ch.9.6.1 for details and examples of usage) to determine approximate value of cost (in the range between 0.1 and 100 -- the suggested range spanning ordes of magnitude should hint that the density of the grid should be approximately logarithmic -- e.g. 1, 3, 10, ... or 1, 2, 5, 10, ... etc.) that yields the lowest error in cross-validation employed by `tune`.  

Setup a resampling procedure repeatedly splitting entire dataset into training and test, using training data to `tune` cost value and test dataset to estimate classification error. Report and discuss distributions of test errors from this procedure and selected values of `cost`.

```{r}
svmfit =svm(auth~., data=dbaDat , kernel ="linear", cost =0.001,scale =TRUE )
summary (svmfit)
```


```{r}
svmfit =svm(auth~., data=dbaDat , kernel ="linear", cost =1,scale =TRUE )
summary (svmfit)
```


```{r}
svmfit =svm(auth~., data=dbaDat , kernel ="linear", cost =1000,scale =TRUE )
summary (svmfit)
```

```{r}
svmfit =svm(auth~., data=dbaDat , kernel ="linear", cost =1000000,scale =TRUE )
summary (svmfit)
```


The higher the cost, the lower no. of support vectors.  

As Victor said on piazza:
"When the cost is 1 million, that the solution for the optimization problem cannot be found within allotted number of iterations."  

In ISLR, it states the reason for that as:
"a smaller value of the cost parameter is being used, we obtain a larger number of support vectors, because the margin is now wider. Unfortunately, the svm() function does not explicitly output the coefficients of
the linear decision boundary obtained when the support vector classifier is fit, nor does it output the width of the margin."  

```{r}
tune.out=tune(svm,auth~.,data=dbaDat,kernel ="linear",ranges =list(cost=c(0.001 , 0.01, 0.1, 1,5,10,100) ))
summary (tune.out)
```

- best parameters:
 cost
    5
    error: 0.008748545
    
```{r}
summary (tune.out$best.model)
```


```{r}
dfOutput = NULL
for (k in c(0.1,1,5,10,25)) {
  trainIdx <- sample(nrow(dbaDat),nrow(dbaDat)/2, replace = TRUE)
    trainD = dbaDat[trainIdx,]
    testD = dbaDat[-trainIdx,]
    tune.out.train=tune(svm , auth ~.,data=trainD ,kernel = "linear",cost= k)
    bestmod.train=tune.out.train$best.model
    ypred=predict(bestmod.train ,testD[,1:4] )
   df = table(predict =ypred , truth= testD$auth)
   
  resCorrect = sum(diag(df))
  resIncorrect = sum(diag(df[nrow(df):1, ]))
  svmerror=1-(resCorrect/sum(df))
  dfOutput = rbind(dfOutput, data.frame(Cost=k, CorrectClassified=resCorrect, Misclassified = resIncorrect,error=svmerror))
}
dfOutput
```

It looks like we have best classification at Cost = 10. There are only 7 misclassified.  

# Problem 2 (10 points): comparison to random forest

Fit random forest classifier on the entire banknote authentication dataset with default parameters.  Calculate resulting misclassification error as reported by the confusion matrix in random forest output.  Explain why error reported in random forest confusion matrix represents estimated test (as opposed to train) error of the procedure.  Compare resulting test error to that for support vector classifier obtained above and discuss results of such comparison.

```{r}
rfRes <- randomForest(auth~.,dbaDat)
rfRes$confusion
sum(diag(rfRes$confusion[nrow(rfRes$confusion):1, ]))
#rfTmpTbl <- table(classTmp[!bTrain],predict(rfRes,newdata=xyzTmp[!bTrain,]))
#rfTmpTbl
nCorrectClass = sum(diag(rfRes$confusion))
misClassificationError = 1 - nCorrectClass/nrow(dbaDat)
misClassificationError
```

The RF performes better than the SVM as it got a smaller error.

# Extra 7 points problem: effect of `mtry` and `ntree` in random forest

Not directly related to SVM, but while we are at it: fit random forest to the entire banknote authentication dataset for every possible value of parameter `mtry` and using `ntree` of 100 and 1000 for each of them.  The values of `mtry` possible in this case are 1, 2, 3 and 4.  Please explain what is governed by this parameter and why this is the exhaustive set of the values allowed for it in this case. Would it change for another dataset?  What is the default value of `mtry` for this dataset?  Repeat this several times to assess center and spread of the error rates produced by random forest with these parameters across multiple runs of random forest procedure.  Present these results graphically and comment on the impact of the choices of `mtry` and `ntree` on the resulting error rates.

```{r}
rfRes <- randomForest(auth~.,dbaDat,mtry=1,ntree=100)
rfRes$confusion
#sum(diag(rfRes$confusion[nrow(rfRes$confusion):1, ]))
plot(rfRes, main = paste0("mtry = ",1,"\n","ntree = ", 100))
```

```{r}
rfRes <- randomForest(auth~.,dbaDat,mtry=2,ntree=100)
rfRes$confusion
plot(rfRes, main = paste0("mtry = ",2,"\n","ntree = ", 100))

```

```{r}
rfRes <- randomForest(auth~.,dbaDat,mtry=3,ntree=100)
rfRes$confusion
plot(rfRes, main = paste0("mtry = ",3,"\n","ntree = ", 100))

```

```{r}
rfRes <- randomForest(auth~.,dbaDat,mtry=4,ntree=100)
rfRes$confusion
plot(rfRes, main = paste0("mtry = ",4,"\n","ntree = ", 100))

```

```{r}
rfRes <- randomForest(auth~.,dbaDat,mtry=1,ntree=1000)
rfRes$confusion
plot(rfRes, main = paste0("mtry = ",1,"\n","ntree = ", 1000))

```

```{r}
rfRes <- randomForest(auth~.,dbaDat,mtry=2,ntree=1000)
rfRes$confusion
plot(rfRes, main = paste0("mtry = ",2,"\n","ntree = ", 1000))

```

```{r}
rfRes <- randomForest(auth~.,dbaDat,mtry=3,ntree=1000)
rfRes$confusion
plot(rfRes, main = paste0("mtry = ",3,"\n","ntree = ", 1000))

```

```{r}
rfRes <- randomForest(auth~.,dbaDat,mtry=4,ntree=1000)
rfRes$confusion
plot(rfRes, main = paste0("mtry = ",4,"\n","ntree = ", 1000))

```

In fact, teh mtry confused me a little because it decreased at the beginning, but with more increase, the error returned to increase.  
Hence, I googled, I found the following from this link:
http://code.env.duke.edu/projects/mget/export/HEAD/MGET/Trunk/PythonPackage/dist/TracOnlineDocumentation/Documentation/ArcGISReference/RandomForestModel.FitToArcGISTable.html  
"mtry did not affect the correct classification rates of their model  
Larger number of trees produce more stable models"     


# Problem 3 (10 points): Comparison to cross-validation tuned KNN predictor

Use convenience wrapper `tune.knn` provided by the library `e1071` on the entire dataset to determine optimal value for the number of the nearest neighbors 'k' to be used in KNN classifier.  Consider our observations in week 9 assignment when choosing range of values of `k` to be evaluated by `tune.knn`.  

Setup resampling procedure similar to that used above for support vector classifier that will repeatedly: 
a) split banknote authentication dataset into training and test, 
b) use `tune.knn` on training data to determine optimal `k`, and 
c) use `k` estimated by `tune.knn` to make KNN classifications on test data.  
Report and discuss distributions of test errors from this procedure and selected values of `k`, 
compare them to those obtained for random forest and support vector classifier above.


```{r}
bestarray=NULL
for(i in 1:10){
trainIdx <- sample(nrow(dbaDat),nrow(dbaDat)/2, replace = TRUE)
    trainD = dbaDat[trainIdx,]
    testD = dbaDat[-trainIdx,]
    knn.cross <- tune.knn(trainD,trainD$auth,k=1:20)
 best.each=knn.cross$best.parameters
 bestarray=rbind(bestarray,best.each$k)
#summary(knn.cross)
}
error=NULL
for(i in 1:nrow(bestarray)){
  knn.res=knn(trainD,testD,factor(trainD$auth),k=bestarray[i])
  knntable=table(knn.res,testD$auth)
  error=rbind(error,data.frame(1-sum(diag(knntable))/sum(knntable)))
  
}
knntable
error

```

Then KNN is the best classifier so far. The error values are mostly zeros except a couple. Then it is better than the linear svm and the random forest.  

# Problem 4 (20 points): SVM with radial kernel

## Sub-problem 4a (10 points): impact of $gamma$ on classification surface

*Plot* SVM model fit to the banknote authentication dataset using (for the ease of plotting) *only variance and skewness* as predictors variables, `kernel="radial"`, `cost=1` and `gamma=1` (see ISLR Ch.9.6.2 for an example of that done with a simulated dataset).  You should be able to see in the resulting plot the magenta-cyan classification boundary as computed by this model.  Produce the same kinds of plots using 0.01 and 100 as values of `gamma` also.  Compare classification boundaries between these three plots and describe how they are impacted by the change in the value of `gamma`.  Can you trace it back to the role of `gamma` in the equation introducing it with the radial kernel in ISLR?

```{r}
dat = data.frame(y=as.factor(dbaDat$auth),x1=dbaDat$var,x2=dbaDat$skew)

svmfit=svm(y~x2+x1,data=dat,kernel ="radial",cost=1,gamma=1)
summary (svmfit)
plot(svmfit,dat)
```


```{r}
dat = data.frame(y=as.factor(dbaDat$auth),x1=dbaDat$var,x2=dbaDat$skew)

svmfit=svm(y~x2+x1,data=dat,kernel ="radial",cost=1,gamma=0.01)
summary (svmfit)
plot(svmfit,dat)

```


```{r}
dat = data.frame(y=as.factor(dbaDat$auth),x1=dbaDat$var,x2=dbaDat$skew)

svmfit=svm(y~x2+x1,data=dat,kernel ="radial",cost=1,gamma=100)
summary (svmfit)
plot(svmfit,dat)

```

The best case is when gamma is the highest, it is almost hugging the correct values, and misclassfied are almost = 0. While when gamma=0.01, the decision boundary is smoother but has many missclassifed points. When gamma=1, the missclassified are less than that of gamma=0.01  


## Sub-problem 4b (10 points): test error for SVM with radial kernel

Similar to how it was done above for support vector classifier (and KNN), set up a resampling process that will repeatedly: 
a) split the entire dataset (using all attributes as predictors) into training and test datasets, 
b) use `tune` function to determine optimal values of `cost` and `gamma` and 
c) calculate test error using these values of `cost` and `gamma`.  
You can start with `cost=c(1,2,5,10,20)` and `gamma=c(0.01,0.02,0.05,0.1,0.2)` as starting ranges to evaluate by `tune`, but please feel free to experiment with different sets of values and discuss the results of it and how you would go about selecting those ranges starting from scratch.  

Present resulting test error graphically, compare it to that of support vector classifier (with linear kernel), random forest and KNN classifiers obtained above and discuss results of these comparisons. 

```{r}
gammalist = NULL
costlist = NULL
dfSVM=NULL
for(i in 1:10){
trainIdx <- sample(nrow(dbaDat),nrow(dbaDat)/2, replace = TRUE)
    trainD = dbaDat[trainIdx,]
    testD = dbaDat[-trainIdx,]
    # knn.cross <- tune.knn(trainD,trainD$auth,k=1:00,tunecontrol=tune.control(sampling = "cross"), cross=10)
    tune.out=tune(svm,auth~.,data=trainD,kernel="radial",ranges=list(cost=c(1,2,5,10,20),gamma=c(0.01,0.02,0.05,0.1,0.2)))
  bestmod.train=tune.out$best.model
   bestg = bestmod.train$gamma
    bestc = bestmod.train$cost
  gammalist = rbind(gammalist,bestg)
  costlist = rbind(costlist,bestc)
  
     ypred=predict(bestmod.train ,testD[,1:4] )
  df = table(predict =ypred , truth= testD$auth)
 tune.error = 1-sum(diag(df))/sum(df)
  dfSVM = rbind(dfSVM, data.frame("Cost" = bestmod.train$cost, "Error" = tune.error))
}  
plot(y=dfSVM$Error,x=dfSVM$Cost,main="Error Rate for different costs")

```

- best parameters:
 cost gamma
   20  0.05
  The higher the cost with modereate gamma, the less error.
  
  
 

```{r}
svmerrordf=data.frame(type="linear",e=dfOutput$error)
randomforesterroe=data.frame(type="RF",e=factor(misClassificationError))
 radialerror=data.frame(cbind(type="radial",e=dfSVM$Error))
 knnerror=data.frame(cbind(type="KNN",e=error$X1...sum.diag.knntable...sum.knntable.))
 errordf=rbind(svmerrordf,radialerror,knnerror,randomforesterroe)
 ggplot(errordf,aes(x=type,y=e))+ geom_boxplot()

```

# Extra 8 points problem: SVM with polynomial kernel

Repeat what was done above (plots of decision boundaries for various interesting values of tuning parameters and test error for their best values estimated from training data) using `kernel="polynomial"`.   Determine ranges of `cost` and `gamma` to be evaluated by `tune`.  Present and discuss resulting test error and how it compares to linear and radial kernels and those of random forest and SVM.

```{r}
dat = data.frame(y=as.factor(dbaDat$auth),x1=dbaDat$var,x2=dbaDat$skew)

svmfit=svm(y~x2+x1,data=dat,kernel ="polynomial",cost=1,gamma=1)
summary (svmfit)
plot(svmfit,dat)
```


```{r}
dat = data.frame(y=as.factor(dbaDat$auth),x1=dbaDat$var,x2=dbaDat$skew)

svmfit=svm(y~x2+x1,data=dat,kernel ="polynomial",cost=1,gamma=0.01)
summary (svmfit)
plot(svmfit,dat)

```


```{r}
dat = data.frame(y=as.factor(dbaDat$auth),x1=dbaDat$var,x2=dbaDat$skew)

svmfit=svm(y~x2+x1,data=dat,kernel ="polynomial",cost=1,gamma=100)
summary (svmfit)
plot(svmfit,dat)

```

The polynomial is performing much worse than the radial. When the gamma =0.01 in polynomial, it could not detect anything, there is no decision boundary, when gamma=100, 


```{r}
set.seed(1)
bTrain <- sample(nrow(dbaDat),nrow(dbaDat)/2, replace = TRUE)
tune.out=tune(svm,auth~.,data=dbaDat[bTrain,],kernel="polynomial",ranges=list(cost=c(1,2,5,10,20),gamma=c(0.01,0.02,0.05,0.1,0.2)))
summary(tune.out)

```

```{r}
dfPoly = NULL
 for (i in 1:30) {
  
   trainIdx <- sample(nrow(dbaDat),nrow(dbaDat)/2, replace = TRUE)
    trainD = dbaDat[trainIdx,]
    testD = dbaDat[-trainIdx,]
    tune.out.train=tune(svm , auth ~.,data=trainD ,kernel = "polynomial",ranges=list(cost=10^(-1:2), gamma=10^(-1:2)))
    bestmod.train=tune.out.train$best.model
  
  #  df=rbind(df,data.frame(bestmod.train$cost,try=i))
    tune.pred.test=predict(bestmod.train,newdata=testD[,1:4])
    
    tmpTbl <- table(truth=testD$auth,predict=tune.pred.test)
     
    tune.error = 1-sum(diag(tmpTbl))/sum(tmpTbl)
    dfPoly = rbind(dfPoly, data.frame("Cost" = bestmod.train$cost, "Error" = tune.error))
 }  
plot(y=dfPoly$Error,x=dfPoly$Cost,main="Error Rate for different costs")
```

```{r}
 polyerror=data.frame(cbind(type="poly",e=dfPoly$Error))

errordf=rbind(svmerrordf,radialerror,knnerror,randomforesterroe,polyerror)
 ggplot(errordf,aes(x=type,y=e))+ geom_boxplot()
```
 
 The KNN is perfomring the best and the one of the radial svm values =zero as well and one of the polynomial=zero also. The linear and the random forest are worse than the rest. One of the linear values is the worest.  
 